Research Experience

### 2010-03-22 doc/thesis/thesis.tex
more work with particle filters.

\subsection*{Introduction to Particle Filters}
Particle filters, also known as Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters. Unlike Markov Chain Monte Carlo estimation, particle
filters extend relatively easily to any finite impulse response (FIR) system.
And unlike Variational Bayes or other gradient descent algorithms, particle
filters can be run online, with relatively lower cost. The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF obviously introduces bias; however, it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose either a set or stream
of measurements $y_t$ are made at discrete times $t = 0, 1, ... $, where
$y_t$ could be a vector of measurements. Then the goal is to find the set
of parameters $\hat{\theta}$ that give the best estimate, $\hat{y}_t$, of $y_t$.
In a biological case, with a model based on first principals, there is 
presumably some true set of parameters, $\theta$, a true model $f$
and a true observation equation, $g$ that dictate how the stimulus 
produce noisy measurements,
\begin{equation}
\dot{x}(t) = f(t, x(t), u(t), \theta) + \nu_d

y(t) = g(t, x(t), u(t), \theta) + \nu_s
\end{equation}

Where $x(t)$ is a vector of state variables, $\theta$ is a vector of system
constants, $\nu_d$ is drift noise (such as a Weiner process), $\nu_s$ is additive noise, $y(t)$ is
the vector of observations. $\nu_d$ may be driven by unknown changes in 
cerebral blood flow, or blood oxygen content, and is assumed to be zero mean.
Similarly $\nu_s$, stationary
noise could be driven by unknown changes due to the heart beat and is also assumed
to be zero mean. There are also non-biological effects that could drive these
noise terms such as inherent MR noise.

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, that $\theta$ is time invariant. This assumption is 
based on the idea that most physiological parameters don't change in the 
short run unless they are driven by some other stimulus. The obvious exception
might be something like the heart; however, period between heartbeats vary
quickly enough that accurately accounting for heart beats necessitates a 
new set of model parameters, or direct outside information. In short we
assume no parameters are time varying, because not information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, suck unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Finally, $x(t)$ will include $\theta$, the unknown model constants, although
obviously that will mean that the vector $\dot{x}$ will always have some members
that are 0. The state space equations are then simplified to:
\begin{equation}
\dot{x}(t) = f(x(t), u(t)) + \nu_d

y(t) = g(x(t)) + \nu_s
\end{equation}

The goal of the particle filter, is then to derive the true value of $x(t)$
for each time, given known inputs $u(t), u(t-1), ... u(0)$  and noise 
measurements produced by the system, $y(t), ... , y(0)$ (every time up to
$t$). To begin with then, the particle filter starts with a prior distribution,
which is used to build a Dirac Mixture Probability Distribution function:

\begin{equation}
P(\hat{x}(0)) = \frac{1}{N}\Sigma_0^{N-1} \delta(x_(0) - \hat{x}_i(0))
\end{equation}

where $\hat{x}(0)$ is an empirical representation of the prior, and $\hat{x}_i(0)$
is the $i^{th}$ particle, a single possible representation of the system at time 0.
Initially, since each $\hat{x}_i$ has been drawn form the prior, and no measurements
have been made, the particles are all equally weighted. When a measurement becomes
available, that information is then incorporated into the posterior by re-weighting
the particles .

@@

\subsection*{My Specific Particle Filte}
Although particle filters are well defined, significant design decisions
are application dependent. The most important choice is the weighting function,
$w(y_t - y(\theta_t))$. Obviously the function should weight $0$ maximally,
the drop off rate is extremely important. A very thin gaussian probability 
distribution function has nice properties, but thin tails. As a result, large
outliers in the measurement vector could easily force all the particles to
have near 0 weights, thus forcing the particle filter to converge improperly. 
On the other end of the spectrum, a cauchy pdf may not
weight particles in the middle enough, preventing the particles to never 
converge. Of course, the scale factor, or variance, also plays a signifcant
role in the convergeance rate. The importance of the weighting function
cannot be overstated, as this is the primary factor in deciding the rate at
which the particle filter converges. As part of the simulations, I compare
a gaussian based weighting function with an exponential weightin function.
In both cases, I scaled the width of the weighting functions automatically based
on the signal variance.

>>>>

\subsection*{Introduction to Particle Filters}
Particle filters, also known as Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, particle
filters extend relatively easily to any finite impulse response (FIR) system.
Unlike Variational Bayes or simple gradient descent algorithms, particle
filters can be run online, with relatively lower cost. The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF obviously introduces bias; however, it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose either a set or stream
of measurements $y_t$ are made at discrete times $t = 0, 1, ... $, where
$y_t$ could be a vector of measurements. Then the goal is to find the set
of parameters $\hat{\theta}$ that give the best estimate, $\hat{y}_t$, of $y_t$.
In a biological case, with a model based on first principals, there is 
presumably some true set of parameters, $\theta$, a true model $f$
and a true observation equation, $g$ that dictate how the stimulus 
produce noisy measurements,
\begin{equation}
\dot{x}(t) = f(t, x(t), u(t), \theta) + \nu_d
\end{equation}

\begin{equation}
y(t) = g(t, x(t), u(t), \theta) + \nu_s
\end{equation}

Where $x(t)$ is a vector of state variables, $\theta$ is a vector of system
constants, $\nu_d$ and $\nu_s$ are bother stationary zero-mean distributions, $u(t)$
is timeseries of stimuli and
$y(t)$ is the vector of observations. $\nu_d$, which produces a drift term in $y(t)$
may be driven by unknown changes in 
cerebral blood flow, or blood oxygen content. $\nu_s$ will produce standard
stationary white noise in $y(t)$ and may encapsulate noise by unknown changes due 
to the heart beat. There are also significant non-biological effects that 
could drive these noise terms such as inherent MR noise.

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, that $\theta$ is time invariant. This assumption is 
based on the idea that most physiological parameters don't change in the 
short run unless they are driven by some other stimulus. The obvious exception
might be something like the heart; however, period between heartbeats vary quickly enough that accurately accounting for them would necessitate an entirely
new set of model parameters, or direct outside information. In short we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so ananticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Finally, $x(t)$ will include $\theta$, the unknown model constants, although
obviously that will mean that the vector $\dot{x}$ will always have some members
that are constant and 0. The state space equations are then simplified to:
\begin{equation}
\dot{x}(t) = f(x(t), u(t)) + \nu_d
\end{equation}

\begin{equation}
y(t) = g(x(t)) + \nu_s
\end{equation}

The goal of the particle filter, is then to derive the probably distribution
of the value of $x(t)$
for each time, given known inputs $u(t), u(t-1), ... u(0)$  and noise 
measurements produced by the system, $y(t), ... , y(0)$ (every time up to
$t$). To begin with then, the particle filter starts with a prior distribution,
which is used to build a Dirac Mixture Probability Distribution Function (PDF):

\begin{equation}
\hat{x}(0) \sim p(X = x) = \frac{1}{n}\sum_{i = 0}^{n-1}p_i(x)
\end{equation}
where $p_i(x)$ is some probability density function centered at $\hat{x}_i(0)$, particle
$i$ at time $0$. Thus the sum of the density at $x$ from each particles' PDF is the density
at $x$. If
the distributions are infinitely thin, then the distribution will be discrete, otherwise
it will be continuous.

Thus $\hat{x}(0)$ is an empirical representation of the prior, and $\hat{x}_i(0)$
is the $i^{th}$ particle, a single possible representation of the system at time 0.
Initially, since each $\hat{x}_i$ has been drawn from the prior, and no measurements
have been made, the particles are all equally weighted. When a measurement becomes
available, that information is then incorporated into the posterior by re-weighting
the particles according to how well they estimated this new measurement. 
\begin{equation}
Pr[\hat{x}_i(t) | y(0:t-1)] = \int Pr[\hat{x}_i(t) | \hat{x}_i(t-1), y(0:t-1)]Pr[\hat{x}_i(t-1)
\end{equation}
Since $\hat{x}(t) = \hat{x}(t-1) + \int_{t-1}^t f(x(t), u(t))dt + \int_{t-1}^t\nu_d dt$, 
\begin{equation}
Pr[\hat{x}_i(t) | \hat{x}_i(t-1)] = Pr[\int{t-1}^t \nu_d]
\end{equation}

of course $\int_{t-1}^t f(x(t), u(t))dt$ may need to be evaluated numerically. Therefore,
to estimate the posterior probability at $t$, it is just necessary to propagate
each particle forward in time according to a state space function. To form 
the posterior probability around the true $x(t)$, each particle is weighted
based on

This
then results in the expected value of $\hat{y}(t)$:
\begin{equation}
E[\hat{y}_i(t) | \hat{x}_i(t-1)] = E[\hat{x}_i(t)]
\end{equation}

@@

\subsection*{My Specific Particle Filte}
Although particle filters are well defined, significant design decisions
are application dependent. The most important choice is the weighting function,
$w(y_t - y(\theta_t))$. Obviously the function should weight $0$ maximally,
the drop off rate is extremely important. A very thin gaussian probability 
distribution function has nice properties, but thin tails. As a result, large
outliers in the measurement vector could easily force all the particles to
have near 0 weights, thus forcing the particle filter to converge improperly. 
On the other end of the spectrum, a cauchy PDF may not
weight particles in the middle enough, preventing the particles to never 
converge. Of course, the scale factor, or variance, also plays a signifcant
role in the convergeance rate. The importance of the weighting function
cannot be overstated, as this is the primary factor in deciding the rate at
which the particle filter converges. As part of the simulations, I compare
a gaussian based weighting function with an exponential weightin function.
In both cases, I scaled the width of the weighting functions automatically based
on the signal variance.
